
```{r setup, include=FALSE}
false=FALSE
true=TRUE
```

#Import
```{r}
library(DESeq2)
library(dplyr)
library(remotes)
library(ggpubr)
library(forcats)
library(ggplot2)
library(stringr)
library(ggrepel)
library(gplots)

countData=read.csv('H:/geo submission/countdata.csv',header=1)
countData_id<-dplyr::select(countData,1,2)
head(countData)
rownames(countData)=countData$gene_symbol
```

#Separate for individual analysis
```{r}
ThreeMonJoint=dplyr::select(countData, 1,2,6,7,8,12,13,14)
rownames(ThreeMonJoint)=ThreeMonJoint$gene_symbol
ThreeMonJoint=ThreeMonJoint[,-1]
ThreeMonJoint=ThreeMonJoint[,-1]

FiveDayJoint=dplyr::select(countData, 1,2,3,4,5,9,10,11)
rownames(FiveDayJoint)=FiveDayJoint$gene_symbol
FiveDayJoint=FiveDayJoint[,-1]
FiveDayJoint=FiveDayJoint[,-1]

WKYdev=dplyr::select(countData, 1,2,3,4,5,6,7,8)
rownames(WKYdev)=WKYdev$gene_symbol
WKYdev=WKYdev[,-1]
WKYdev=WKYdev[,-1]

SHRdev=dplyr::select(countData, 1,2,9,10,11,12,13,14)
rownames(SHRdev)=SHRdev$gene_symbol
SHRdev=SHRdev[,-1]
SHRdev=SHRdev[,-1]

countData=countData[,-1]
countData=countData[,-1]

strain=c("WKY","WKY","WKY","WKY","WKY","WKY","SHR","SHR","SHR","SHR","SHR","SHR")
cellType=c("5d_Chon","5d_Chon","5d_Chon","3m_Chon","3m_Chon","3m_Chon","5d_Chon","5d_Chon","5d_Chon","3m_Chon","3m_Chon","3m_Chon")
batch=c("1","1","2","1","1","2","1","1","2","1","1","2")

strainByType=c("WKY","WKY","WKY","SHR","SHR","SHR")
batchByType=c("1","1","2","1","1","2")
cellByTime=c("5d","5d","5d","3m","3m","3m")

colData=as.data.frame(cbind(colnames(countData),strain,cellType,batch))
colData3m=as.data.frame(cbind(colnames(ThreeMonJoint),strainByType,batchByType))
colData5d=as.data.frame(cbind(colnames(FiveDayJoint),strainByType,batchByType))
colDataWKYdev=as.data.frame(cbind(colnames(WKYdev),cellByTime,batchByType))
colDataSHRdev=as.data.frame(cbind(colnames(SHRdev),cellByTime,batchByType))
```

#Differential analysis
```{r}
countData=trunc(countData)
dds=DESeqDataSetFromMatrix(countData=countData,colData=colData,design=~strain+cellType+strain*cellType+batch)

ThreeMonJoint=trunc(ThreeMonJoint)
FiveDayJoint=trunc(FiveDayJoint)
WKYdev=trunc(WKYdev)
SHRdev=trunc(SHRdev)
dds3m=DESeqDataSetFromMatrix(countData=ThreeMonJoint,colData=colData3m,design=~strainByType+batchByType)
dds5d=DESeqDataSetFromMatrix(countData=FiveDayJoint,colData=colData5d,design=~strainByType+batchByType)
ddsWKY=DESeqDataSetFromMatrix(countData=WKYdev,colData=colDataWKYdev,design=~cellByTime+batchByType)
ddsSHR=DESeqDataSetFromMatrix(countData=SHRdev,colData=colDataSHRdev,design=~cellByTime+batchByType)

##remove batch effect

dds <- DESeq(dds, test=c("LRT"),full=design(dds), reduced = ~ batch)
dds3m <- DESeq(dds3m, test=c("LRT"),full=design(dds3m), reduced = ~ batchByType)
dds5d <- DESeq(dds5d, test=c("LRT"),full=design(dds5d), reduced = ~ batchByType)
ddsWKY <- DESeq(ddsWKY, test=c("LRT"),full=design(ddsWKY), reduced = ~ batchByType)
ddsSHR <- DESeq(ddsSHR, test=c("LRT"),full=design(ddsSHR), reduced = ~ batchByType)
```

#Normalized expression
```{r}
rld3m=rlog(dds3m)
mat <- assay(rld3m)
mm <- model.matrix(~strainByType, colData(rld3m))
mat <- limma::removeBatchEffect(mat, batch=rld3m$batchByType, design=mm)
assay(rld3m) <- mat

rld5d=rlog(dds5d)
mat <- assay(rld5d)
mm <- model.matrix(~strainByType, colData(rld5d))
mat <- limma::removeBatchEffect(mat, batch=rld5d$batchByType, design=mm)
assay(rld5d) <- mat

rldWKY=rlog(ddsWKY)
mat <- assay(rldWKY)
mm <- model.matrix(~cellByTime, colData(rldWKY))
mat <- limma::removeBatchEffect(mat, batch=rldWKY$batchByType, design=mm)
assay(rldWKY) <- mat

rldSHR=rlog(ddsSHR)
mat <- assay(rldSHR)
mm <- model.matrix(~cellByTime, colData(rldSHR))
mat <- limma::removeBatchEffect(mat, batch=rldSHR$batchByType, design=mm)
assay(rldSHR) <- mat

rldChon=rlog(dds)
mat <- assay(rldChon)
mm <- model.matrix(~strain+cellType+strain*cellType, colData(rldChon))
mat <- limma::removeBatchEffect(mat, batch=rldChon$batch, design=mm)
assay(rldChon) <- mat
```

##Principal component analysis
```{r}
library(ggplot2)
library(ggforce)

plotPCA(rldChon,intgroup=c("strain","cellType")) + geom_point(aes(size = 2),show.legend=F) + geom_mark_ellipse(aes(fill = group), alpha = 0.25)
```

#Comparison results
```{r}
res3m=results(dds3m,contrast=c("strainByType","SHR","WKY"),alpha=0.05)
res3m=res3m[order(abs(res3m$pvalue),decreasing=FALSE),]
head(res3m)
summary(res3m)

res5d=results(dds5d,contrast=c("strainByType","SHR","WKY"),alpha=0.05)
res5d=res5d[order(abs(res5d$pvalue),decreasing=FALSE),]
head(res5d)
summary(res5d)

resWKY=results(ddsWKY,contrast=c("cellByTime","3m","5d"),alpha=0.05)
resWKY=resWKY[order(abs(resWKY$pvalue),decreasing=FALSE),]
head(resWKY)
summary(resWKY)

resSHR=results(ddsSHR,contrast=c("cellByTime","3m","5d"),alpha=0.05)
resSHR=resSHR[order(abs(resSHR$pvalue),decreasing=FALSE),]
head(resSHR)
summary(resSHR)
```

#Volcano plots
```{r}
library(dplyr)
library(ggplot2)
res3m1=as.data.frame(res3m)

res3m1$geneid<-rownames(res3m1)
res3m1=mutate(res3m1,sig="No change")
res3m1$sig[res3m1$log2FoldChange > 1 & res3m1$padj < 0.05] <- "Upregulated"
res3m1$sig[res3m1$log2FoldChange < -1 & res3m1$padj < 0.05] <- "Downregulated"

# Identify the top 10 hits
Top_Hits = subset(res3m1,res3m1$padj<(1e-15) | abs(res3m1$log2FoldChange) > 7)
Top_Hits = subset(Top_Hits,abs(Top_Hits$log2FoldChange)>1 & Top_Hits$padj<0.05)

# Add column label, containing the gene name for the top hits or nothing for all others
res3m1$label = if_else(res3m1$geneid %in% Top_Hits$geneid,  
                   res3m1$geneid, "")

ggplot(data=res3m1, aes(x=log2FoldChange, y=-log10(padj), col=sig, label=label)) +
        geom_point() + 
          scale_color_manual(values=c("#619CFF","#00BA38","#F8766D"))+
        geom_vline(xintercept=c(-1, 1), col="red",linetype = "dashed") +
        geom_hline(yintercept=-log10(0.05), col="red",linetype = "dashed") + 
  ggtitle("3-month-old Joint Chondrocytes: WKY vs SHR")+
  labs(color="Significantly Different?")+ 
  geom_text_repel(check_overlap = TRUE,show.legend = FALSE,fontface = "bold",size=2)
```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
res5d1=as.data.frame(res5d)

res5d1=mutate(res5d1,sig="No change")
res5d1$sig[res5d1$log2FoldChange > 1 & res5d1$padj < 0.05] <- "Upregulated"
res5d1$sig[res5d1$log2FoldChange < -1 & res5d1$padj < 0.05] <- "Downregulated"

res5d1$geneid<-rownames(res5d1)
# Identify the top 10 hits
Top_Hits = subset(res5d1,res5d1$padj<(1e-15) | abs(res5d1$log2FoldChange) > 7)
Top_Hits = subset(Top_Hits,abs(Top_Hits$log2FoldChange)>1 & Top_Hits$padj<0.05)

# Add column label, containing the gene name for the top hits or nothing for all others
res5d1$label = if_else(res5d1$geneid %in% Top_Hits$geneid,  
                   res5d1$geneid, "")

ggplot(data=res5d1, aes(x=log2FoldChange, y=-log10(padj), col=sig,label=label)) +
        geom_point() + 
          scale_color_manual(values=c("#619CFF","#00BA38","#F8766D"))+
        geom_vline(xintercept=c(-1, 1), col="red",linetype = "dashed") +
        geom_hline(yintercept=-log10(0.05), col="red",linetype = "dashed") + 
  ggtitle("5-day-old Joint Chondrocytes: WKY vs SHR")+
  labs(color="Significantly Different?")+ 
    geom_text_repel(check_overlap = TRUE,show.legend = FALSE,fontface = "bold",size=2)
```

```{r}
library(dplyr)
resWKY1=as.data.frame(resWKY)

resWKY1=mutate(resWKY1,sig="No change")
resWKY1$sig[resWKY1$log2FoldChange > 1 & resWKY1$padj < 0.05] <- "Upregulated"
resWKY1$sig[resWKY1$log2FoldChange < -1 & resWKY1$padj < 0.05] <- "Downregulated"

resWKY1$geneid<-rownames(resWKY1)
# Identify the top 10 hits
Top_Hits = subset(resWKY1,resWKY1$padj<(1e-15) | abs(resWKY1$log2FoldChange) > 7)
Top_Hits = subset(Top_Hits,abs(Top_Hits$log2FoldChange)>1 & Top_Hits$padj<0.05)

# Add column label, containing the gene name for the top hits or nothing for all others
resWKY1$label = if_else(resWKY1$geneid %in% Top_Hits$geneid,  
                   resWKY1$geneid, "")

ggplot(data=resWKY1, aes(x=log2FoldChange, y=-log10(padj), col=sig,label=label)) +
        geom_point() + 
          scale_color_manual(values=c("#619CFF","#00BA38","#F8766D"))+
        geom_vline(xintercept=c(-1, 1), col="red",linetype = "dashed") +
        geom_hline(yintercept=-log10(0.05), col="red",linetype = "dashed") + 
  ggtitle("WKY: Joint_Chondrocytes_5-day-old vs 3-month-old")+labs(color="Significantly Different?")+ 
  geom_text_repel(check_overlap = TRUE,show.legend = FALSE,fontface = "bold",size=2)
```

```{r}
library(dplyr)
resSHR1=as.data.frame(resSHR)

resSHR1=mutate(resSHR1,sig="No change")
resSHR1$sig[resSHR1$log2FoldChange > 1 & resSHR1$padj < 0.05] <- "Upregulated"
resSHR1$sig[resSHR1$log2FoldChange < -1 & resSHR1$padj < 0.05] <- "Downregulated"

resSHR1$geneid<-rownames(resSHR1)
# Identify the top 10 hits
Top_Hits = subset(resSHR1,resSHR1$padj<(1e-15) | abs(resSHR1$log2FoldChange) > 7)
Top_Hits = subset(Top_Hits,abs(Top_Hits$log2FoldChange)>1 & Top_Hits$padj<0.05)

# Add column label, containing the gene name for the top hits or nothing for all others
resSHR1$label = if_else(resSHR1$geneid %in% Top_Hits$geneid,  
                   resSHR1$geneid, "")

ggplot(data=resSHR1, aes(x=log2FoldChange, y=-log10(padj), col=sig,label=label)) +
        geom_point() + 
          scale_color_manual(values=c("#619CFF","#00BA38","#F8766D"))+
        geom_vline(xintercept=c(-1, 1), col="red",linetype = "dashed") +
        geom_hline(yintercept=-log10(0.05), col="red",linetype = "dashed") + 
  ggtitle("SHR: Joint_Chondrocytes_5-day-old vs 3-month-old")+labs(color="Significantly Different?")+ 
  geom_text_repel(check_overlap = TRUE,show.legend = FALSE,fontface = "bold",size=2)
```


#KEGG SETUP
```{r}
library(pathview)
library(gage)
library(gageData)
library(AnnotationDbi)
library(org.Rn.eg.db)
library(stringr)

data("go.sets.rn")
data("go.subs.rn")
data(kegg.sets.rn)
data(sigmet.idx.rn)

keggData<- kegg.gsets(species = "rno", id.type = "entrez", check.new=T)
kegg.sets.rn<- keggData$kg.sets

#calculate rich ratio
totalGeneSet=nchar(kegg.sets.rn)
totalGeneSet=as.data.frame(totalGeneSet)
totalGeneSet$GeneNoTotal=stringi::stri_count_boundaries(kegg.sets.rn)
totalGeneSet$id=row.names(totalGeneSet)
```

##KEGG general pathways
```{r}
library(dplyr)
res3m2<-as.data.frame(res3m1)
res3m2$gene_symbol<-rownames(res3m2)
res3m3<-inner_join(res3m2,countData_id,by="gene_symbol")

res3m5=as.data.frame(subset(res3m3,pvalue<0.05))
FC_3m=res3m5$log2FoldChange

names(FC_3m)=res3m5$gene_id
head(FC_3m)

kegg3m=gage(exprs=FC_3m,gsets=kegg.sets.rn,same.dir=TRUE)
kegg3mUp1=kegg3m$greater
kegg3mDown1=kegg3m$less
```

```{r}
res5d2<-as.data.frame(res5d1)
res5d2$gene_symbol<-rownames(res5d2)
res5d3<-inner_join(res5d2,countData_id,by="gene_symbol")

res5d5=as.data.frame(subset(res5d3,pvalue<0.05))
FC_5d=res5d5$log2FoldChange

names(FC_5d)=res5d5$gene_id
head(FC_5d)

kegg5d=gage(exprs=FC_5d,gsets=kegg.sets.rn,same.dir=TRUE)
kegg5dUp1=kegg5d$greater
kegg5dDown1=kegg5d$less
```

```{r}
resWKY2<-as.data.frame(resWKY1)
resWKY2$gene_symbol<-rownames(resWKY2)
resWKY3<-inner_join(resWKY2,countData_id,by="gene_symbol")

resWKY5=as.data.frame(subset(resWKY3,pvalue<0.05))
FC_WKY=resWKY5$log2FoldChange

names(FC_WKY)=resWKY5$gene_id
head(FC_WKY)

keggWKY=gage(exprs=FC_WKY,gsets=kegg.sets.rn,same.dir=TRUE)
keggWKYUp1=keggWKY$greater
keggWKYDown1=keggWKY$less
```

```{r}
resSHR2<-as.data.frame(resSHR1)
resSHR2$gene_symbol<-rownames(resSHR2)
resSHR3<-inner_join(resSHR2,countData_id,by="gene_symbol")

resSHR5=as.data.frame(subset(resSHR3,pvalue<0.05))
FC_SHR=resSHR5$log2FoldChange

names(FC_SHR)=resSHR5$gene_id
head(FC_SHR)

keggSHR=gage(exprs=FC_SHR,gsets=kegg.sets.rn,same.dir=TRUE)
keggSHRUp1=keggSHR$greater
keggSHRDown1=keggSHR$less
```

```{r}
library(ggplot2)
kegg3mUp2<-as.data.frame(kegg3mUp1[1:20,])
kegg3mUp2<-subset(kegg3mUp2,p.val>0)
kegg3mUp2$pathway=row.names(kegg3mUp2)

richRatio3mUp=as.data.frame(cbind(kegg3mUp2[,5],kegg3mUp2[,7]))
colnames(richRatio3mUp)=c("GeneNumber","id")
richRatio3mUp1=inner_join(x = totalGeneSet, y = richRatio3mUp, by = 'id')
kegg3mUp2$ratio3m=as.integer(richRatio3mUp1$GeneNumber)/as.integer(richRatio3mUp1$GeneNoTotal)

kegg3mUp2$pathway <- factor(kegg3mUp2$pathway, levels = kegg3mUp2$pathway[order(-(kegg3mUp2$p.val))])

ggplot(kegg3mUp2, aes(x=ratio3m, y=pathway, size=set.size, color=p.val)) + geom_point() + 
theme_classic() +scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab",  limits = c(0, 0.08))+
  xlab("Rich Ratio")+labs(color="P value",size="Gene Number")+ggtitle("3-month-old Joint: WKY vs SHR")+
  ylab("Upregulated KEGG Pathways")+theme(aspect.ratio=1.5)

kegg3mDown2<-as.data.frame(kegg3mDown1[1:20,])
kegg3mDown2<-subset(kegg3mDown2,p.val>0)
kegg3mDown2$pathway=row.names(kegg3mDown2)

richRatio3mDown=as.data.frame(cbind(kegg3mDown2[,5],kegg3mDown2[,7]))
colnames(richRatio3mDown)=c("GeneNumber","id")
richRatio3mDown1=inner_join(x = totalGeneSet, y = richRatio3mDown, by = 'id')
kegg3mDown2$ratio3m=as.integer(richRatio3mDown1$GeneNumber)/as.integer(richRatio3mDown1$GeneNoTotal)

kegg3mDown2$pathway <- factor(kegg3mDown2$pathway, levels = kegg3mDown2$pathway[order(-(kegg3mDown2$p.val))])

ggplot(kegg3mDown2, aes(x=ratio3m, y=pathway, size=set.size, color=p.val)) + geom_point() + 
theme_classic() +scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab")+xlab("Rich Ratio")+labs(color="P value",size="Gene Number")+ggtitle("3-month-old Joint: WKY vs SHR")+ylab("Downregulated KEGG Pathways")+theme(aspect.ratio=1.5)
```

```{r}
library(ggplot2)

kegg5dUp2<-as.data.frame(kegg5dUp1[1:20,])
kegg5dUp2<-subset(kegg5dUp2,p.val>0)
kegg5dUp2$pathway=row.names(kegg5dUp2)

richRatio5dUp=as.data.frame(cbind(kegg5dUp2[,5],kegg5dUp2[,7]))
colnames(richRatio5dUp)=c("GeneNumber","id")
richRatio5dUp1=inner_join(x = totalGeneSet, y = richRatio5dUp, by = 'id')
kegg5dUp2$ratio5d=as.integer(richRatio5dUp1$GeneNumber)/as.integer(richRatio5dUp1$GeneNoTotal)

kegg5dUp2$pathway <- factor(kegg5dUp2$pathway, levels = kegg5dUp2$pathway[order(-(kegg5dUp2$p.val))])

ggplot(kegg5dUp2, aes(x=ratio5d, y=pathway, size=set.size, color=p.val)) + geom_point() + 
theme_classic() +scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab",limits = c(0, 0.2))+
  xlab("Rich Ratio")+labs(color="P value",size="Gene Number")+ggtitle("5-day-old Joint: WKY vs SHR")+
  ylab("Upregulated KEGG Pathways")+theme(aspect.ratio=1.5)

kegg5dDown2<-as.data.frame(kegg5dDown1[1:20,])
kegg5dDown2<-subset(kegg5dDown2,p.val>0)
kegg5dDown2$pathway=row.names(kegg5dDown2)

richRatio5dDown=as.data.frame(cbind(kegg5dDown2[,5],kegg5dDown2[,7]))
colnames(richRatio5dDown)=c("GeneNumber","id")
richRatio5dDown1=inner_join(x = totalGeneSet, y = richRatio5dDown, by = 'id')
kegg5dDown2$ratio5d=as.integer(richRatio5dDown1$GeneNumber)/as.integer(richRatio5dDown1$GeneNoTotal)

kegg5dDown2$pathway <- factor(kegg5dDown2$pathway, levels = kegg5dDown2$pathway[order(-(kegg5dDown2$p.val))])

ggplot(kegg5dDown2, aes(x=ratio5d, y=pathway, size=set.size, color=p.val)) + geom_point() + 
theme_classic() +scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab")+xlab("Rich Ratio")+labs(color="P value",size="Gene Number")+ggtitle("5-day-old Joint: WKY vs SHR")+ylab("Downregulated KEGG Pathways")+theme(aspect.ratio=1.5)
```

```{r}
library(ggplot2)

keggWKYUp2<-as.data.frame(keggWKYUp1[1:20,])
keggWKYUp2<-subset(keggWKYUp2,p.val>0)
keggWKYUp2$pathway=row.names(keggWKYUp2)

richRatioWKYUp=as.data.frame(cbind(keggWKYUp2[,5],keggWKYUp2[,7]))
colnames(richRatioWKYUp)=c("GeneNumber","id")
richRatioWKYUp1=inner_join(x = totalGeneSet, y = richRatioWKYUp, by = 'id')
keggWKYUp2$ratioWKY=as.integer(richRatioWKYUp1$GeneNumber)/as.integer(richRatioWKYUp1$GeneNoTotal)

keggWKYUp2$pathway <- factor(keggWKYUp2$pathway, levels = keggWKYUp2$pathway[order(-(keggWKYUp2$p.val))])

ggplot(keggWKYUp2, aes(x=ratioWKY, y=pathway, size=set.size, color=p.val)) + geom_point() + 
theme_classic() +scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab",limits = c(0, 0.16))+
  xlab("Rich Ratio")+labs(color="P value",size="Gene Number")+
  ggtitle("WKY Joint Chon: 5-day-old vs 3-month-old")+ylab("Upregulated KEGG Pathways")+ 
    theme(plot.title = element_text(size = 9, face = "bold"))

keggWKYDown2<-as.data.frame(keggWKYDown1[1:20,])
keggWKYDown2<-subset(keggWKYDown2,p.val>0)
keggWKYDown2$pathway=row.names(keggWKYDown2)

richRatioWKYDown=as.data.frame(cbind(keggWKYDown2[,5],keggWKYDown2[,7]))
colnames(richRatioWKYDown)=c("GeneNumber","id")
richRatioWKYDown1=inner_join(x = totalGeneSet, y = richRatioWKYDown, by = 'id')
keggWKYDown2$ratioWKY=as.integer(richRatioWKYDown1$GeneNumber)/as.integer(richRatioWKYDown1$GeneNoTotal)

keggWKYDown2$pathway <- factor(keggWKYDown2$pathway, levels = keggWKYDown2$pathway[order(-(keggWKYDown2$p.val))])

ggplot(keggWKYDown2, aes(x=ratioWKY, y=pathway, size=set.size, color=p.val)) + geom_point() + 
theme_classic() +scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab")+xlab("Rich Ratio")+labs(color="P value",size="Gene Number")+
  ggtitle("WKY Joint Chon: 5-day-old vs 3-month-old")+ylab("Downregulated KEGG Pathways")+ 
    theme(plot.title = element_text(size = 9, face = "bold"))
```

```{r}
library(ggplot2)

keggSHRUp2<-as.data.frame(keggSHRUp1[1:20,])
keggSHRUp2<-subset(keggSHRUp2,p.val>0)
keggSHRUp2$pathway=row.names(keggSHRUp2)

richRatioSHRUp=as.data.frame(cbind(keggSHRUp2[,5],keggSHRUp2[,7]))
colnames(richRatioSHRUp)=c("GeneNumber","id")
richRatioSHRUp1=inner_join(x = totalGeneSet, y = richRatioSHRUp, by = 'id')
keggSHRUp2$ratioSHR=as.integer(richRatioSHRUp1$GeneNumber)/as.integer(richRatioSHRUp1$GeneNoTotal)

keggSHRUp2$pathway <- factor(keggSHRUp2$pathway, levels = keggSHRUp2$pathway[order(-(keggSHRUp2$p.val))])

ggplot(keggSHRUp2, aes(x=ratioSHR, y=pathway, size=set.size, color=p.val)) + geom_point() + 
theme_classic() +scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab", limits = c(0, 0.015))+
  xlab("Rich Ratio")+labs(color="P value",size="Gene Number")+
  ggtitle("SHR Joint Chon: 5-day-old vs 3-month-old")+ylab("Upregulated KEGG Pathways")+
    theme(plot.title = element_text(size = 9, face = "bold")) 

keggSHRDown2<-as.data.frame(keggSHRDown1[1:20,])
keggSHRDown2<-subset(keggSHRDown2,p.val>0)
                       
keggSHRDown2$pathway=row.names(keggSHRDown2)

richRatioSHRDown=as.data.frame(cbind(keggSHRDown2[,5],keggSHRDown2[,7]))
colnames(richRatioSHRDown)=c("GeneNumber","id")
richRatioSHRDown1=inner_join(x = totalGeneSet, y = richRatioSHRDown, by = 'id')
keggSHRDown2$ratioSHR=as.integer(richRatioSHRDown1$GeneNumber)/as.integer(richRatioSHRDown1$GeneNoTotal)

keggSHRDown2$pathway <- factor(keggSHRDown2$pathway, levels = keggSHRDown2$pathway[order(-(keggSHRDown2$p.val))])

ggplot(keggSHRDown2, aes(x=ratioSHR, y=pathway, size=set.size, color=p.val)) + geom_point() + 
theme_classic() +scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab")+xlab("Rich Ratio")+
  labs(color="P value",size="Gene Number")+
  ggtitle("SHR Joint Chon: 5-day-old vs 3-month-old")+ylab("Downregulated KEGG Pathways") + 
    theme(plot.title = element_text(size = 9, face = "bold"))
```

##KEGG metabolic pathways
```{r}
kegg.sets.rn.met=list()
for(i in 1:length(kegg.sets.rn)){
  if(str_detect(names(kegg.sets.rn[i]), "rno00") || str_detect(names(kegg.sets.rn[i]), "rno01") & !str_detect(names(kegg.sets.rn[i]), "rno01100"))
  {kegg.sets.rn.met[i]<-kegg.sets.rn[i]
  names(kegg.sets.rn.met)[i]<-names(kegg.sets.rn)[i]
  }
}

#metabolism pathways only
kegg.sets.rn<-kegg.sets.rn.met

#calculate rich ratio
totalGeneSet=nchar(kegg.sets.rn)
totalGeneSet=as.data.frame(totalGeneSet)
totalGeneSet$GeneNoTotal=stringi::stri_count_boundaries(kegg.sets.rn)
totalGeneSet$id=row.names(totalGeneSet)
```

```{r}
resWKY2<-as.data.frame(resWKY1)
resWKY2$gene_symbol<-rownames(resWKY2)
resWKY3<-inner_join(resWKY2,countData_id,by="gene_symbol")

resWKY5=as.data.frame(subset(resWKY3,pvalue<0.05))
FC_WKY=resWKY5$log2FoldChange

names(FC_WKY)=resWKY5$gene_id
head(FC_WKY)

keggWKY=gage(exprs=FC_WKY,gsets=kegg.sets.rn,same.dir=TRUE)
keggWKYUp1=keggWKY$greater
keggWKYDown1=keggWKY$less
```

```{r}
resSHR2<-as.data.frame(resSHR1)
resSHR2$gene_symbol<-rownames(resSHR2)
resSHR3<-inner_join(resSHR2,countData_id,by="gene_symbol")

resSHR5=as.data.frame(subset(resSHR3,pvalue<0.05))
FC_SHR=resSHR5$log2FoldChange

names(FC_SHR)=resSHR5$gene_id
head(FC_SHR)

keggSHR=gage(exprs=FC_SHR,gsets=kegg.sets.rn,same.dir=TRUE)
keggSHRUp1=keggSHR$greater
keggSHRDown1=keggSHR$less
```

```{r}
#circular plot
keggSHRUp2<-as.data.frame(keggSHRUp1[1:10,])
keggSHRUp2<-subset(keggSHRUp2,p.val>0)
keggSHRUp2$pathway=row.names(keggSHRUp2)

richRatioSHRUp=as.data.frame(cbind(keggSHRUp2[,5],keggSHRUp2[,7]))
colnames(richRatioSHRUp)=c("GeneNumber","id")
richRatioSHRUp1=inner_join(x = totalGeneSet, y = richRatioSHRUp, by = 'id')
keggSHRUp2$ratioSHR=as.integer(richRatioSHRUp1$GeneNumber)/as.integer(richRatioSHRUp1$GeneNoTotal)

keggSHRUp2$pathway <- factor(keggSHRUp2$pathway, levels = keggSHRUp2$pathway[order(-(keggSHRUp2$p.val))])

p<-ggplot(keggSHRUp2) + theme_minimal() + 
  geom_hline(aes(yintercept = y), data.frame(y = c(0:5) * 8), color = "lightgrey") + 
  geom_col(aes(x=fct_inorder(str_wrap(pathway,10)), y=ratioSHR*100, size=ratioSHR*100, fill=p.val, alpha=.9)) +   
  scale_fill_gradient(low = "red2",  high = "mediumblue", space = "Lab") +  
  theme(axis.title.y = element_blank(), axis.ticks = element_blank(), axis.text.y = element_blank(), axis.text.x = element_blank())+
  labs(fill="P value")+ guides(size="none",alpha="none") +
  ggtitle("SHR Joint Chon: 5-day-old vs 3-month-old")+xlab("Upregulated KEGG Pathways") + 
    theme(plot.title = element_text(size = 12, face = "bold",vjust=-15)) +  coord_polar()+
    annotate(x = 0.5,  y = 8, label = "8%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
    annotate(x = 0.5,  y = 16, label = "16%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
    annotate(x = 0.5,  y = 24, label = "24%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
    annotate(x = 0.5,  y = 32, label = "32%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
    annotate(x = 0.5,  y = 40, label = "40%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
   annotate(x = 0.5,  y = 45, label = "Rich Ratio",  geom = "text", color = "gray12", size=4, vjust=0.5, fontface = "bold")+
   theme(plot.background = element_rect(fill = "white",colour = NA_character_))+
 # theme(axis.text.x = element_text( face="bold", size=12))+ 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
    theme(axis.title.x = element_text(size = 12, face = "bold",vjust=25))

for (i in 1:10){
  p <- p+ annotate(x = i,  y = keggSHRUp2$ratioSHR[i]*100+10, label = str_wrap(keggSHRUp2$pathway[i],10),  geom = "text", 
                   color = "gray12", size=3, vjust=0.5, fontface = "bold")
}
  p
```

```{r}
#circular plot
keggWKYUp2<-as.data.frame(keggWKYUp1[1:10,])
keggWKYUp2<-subset(keggWKYUp2,p.val>0)
keggWKYUp2$pathway=row.names(keggWKYUp2)

richRatioWKYUp=as.data.frame(cbind(keggWKYUp2[,5],keggWKYUp2[,7]))
colnames(richRatioWKYUp)=c("GeneNumber","id")
richRatioWKYUp1=inner_join(x = totalGeneSet, y = richRatioWKYUp, by = 'id')
keggWKYUp2$ratioWKY=as.integer(richRatioWKYUp1$GeneNumber)/as.integer(richRatioWKYUp1$GeneNoTotal)

keggWKYUp2$pathway <- factor(keggWKYUp2$pathway, levels = keggWKYUp2$pathway[order(-(keggWKYUp2$p.val))])

p<-ggplot(keggWKYUp2) + theme_minimal() + 
  geom_hline(aes(yintercept = y), data.frame(y = c(0:5) * 8), color = "lightgrey") + 
  geom_col(aes(x=fct_inorder(str_wrap(pathway,10)), y=ratioWKY*100, size=ratioWKY*100, fill=p.val, alpha=.9)) +   
  scale_fill_gradient(low = "red2",  high = "mediumblue", space = "Lab") +  
  theme(axis.title.y = element_blank(), axis.ticks = element_blank(), axis.text.y = element_blank(), axis.text.x = element_blank())+
  labs(fill="P value")+ guides(size="none",alpha="none") +
  ggtitle("WKY Joint Chon: 5-day-old vs 3-month-old")+xlab("Upregulated KEGG Pathways") + 
    theme(plot.title = element_text(size = 12, face = "bold",vjust=-15)) +  coord_polar()+
    annotate(x = 0.5,  y = 8, label = "8%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
    annotate(x = 0.5,  y = 16, label = "16%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
    annotate(x = 0.5,  y = 24, label = "24%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
    annotate(x = 0.5,  y = 32, label = "32%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
    annotate(x = 0.5,  y = 40, label = "40%",  geom = "text", color = "gray12", size=3.5, vjust=0.5, fontface = "bold")+
   annotate(x = 0.5,  y = 45, label = "Rich Ratio",  geom = "text", color = "gray12", size=4, vjust=0.5, fontface = "bold")+
   theme(plot.background = element_rect(fill = "white",colour = NA_character_))+
 # theme(axis.text.x = element_text( face="bold", size=12))+ 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
    theme(axis.title.x = element_text(size = 12, face = "bold",vjust=25))

for (i in 1:10){
  p <- p+ annotate(x = i,  y = keggWKYUp2$ratioWKY[i]*100+10, label = str_wrap(keggWKYUp2$pathway[i],10),  geom = "text", 
                   color = "gray12", size=3, vjust=0.5, fontface = "bold")
}
  p
```

#Gene Ontology Analysis
```{r}
library(GO.db)
library(GOstats)
```

##GO individual codes
```{r}
#3m
res3m$id<-rownames(res3m)
res3m$entrez<-mapIds(org.Rn.eg.db,key=res3m$id,column="ENTREZID",keytype="SYMBOL")
res3m5=as.data.frame(subset(res3m,padj<0.05))
universeGenes=res3m$entrez

selectGenesUp3m=subset(res3m5,res3m5$log2FoldChange>0)$entrez
selectGenesDown3m=subset(res3m5,res3m5$log2FoldChange<0)$entrez

upParams3m=new("GOHyperGParams",geneIds=selectGenesUp3m,universeGeneIds=universeGenes,annotation="org.Rn.eg.db",ontology="BP",pvalueCutoff=0.05, conditional=FALSE,testDirection="over")
downParams3m=new("GOHyperGParams",geneIds=selectGenesDown3m,universeGeneIds=universeGenes,annotation="org.Rn.eg.db",ontology="BP",pvalueCutoff=0.05, conditional=FALSE,testDirection="over")

upBP_3m=hyperGTest(upParams3m)
upBP1_3m=summary(upBP_3m)
downBP_3m=hyperGTest(downParams3m)
downBP1_3m=summary(downBP_3m)

##
ontology(upParams3m)="CC"
upCC_3m=hyperGTest(upParams3m)
upCC1_3m=summary(upCC_3m)

ontology(downParams3m)="CC"
downCC_3m=hyperGTest(downParams3m)
downCC1_3m=summary(downCC_3m)

##
ontology(upParams3m)="MF"
upMF_3m=hyperGTest(upParams3m)
upMF1_3m=summary(upMF_3m)

ontology(downParams3m)="MF"
downMF_3m=hyperGTest(downParams3m)
downMF1_3m=summary(downMF_3m)
```

```{r}
#5d
res5d$id<-rownames(res5d)
res5d$entrez<-mapIds(org.Rn.eg.db,key=res5d$id,column="ENTREZID",keytype="SYMBOL")
res5d5=as.data.frame(subset(res5d,padj<0.05))
universeGenes=res5d$entrez

selectGenesUp5d=subset(res5d5,res5d5$log2FoldChange>0)$entrez
selectGenesDown5d=subset(res5d5,res5d5$log2FoldChange<0)$entrez

upParams5d=new("GOHyperGParams",geneIds=selectGenesUp5d,universeGeneIds=universeGenes,annotation="org.Rn.eg.db",ontology="BP",pvalueCutoff=0.05, conditional=FALSE,testDirection="over")
downParams5d=new("GOHyperGParams",geneIds=selectGenesDown5d,universeGeneIds=universeGenes,annotation="org.Rn.eg.db",ontology="BP",pvalueCutoff=0.05, conditional=FALSE,testDirection="over")

upBP_5d=hyperGTest(upParams5d)
upBP1_5d=summary(upBP_5d)

downBP_5d=hyperGTest(downParams5d)
downBP1_5d=summary(downBP_5d)

##
ontology(upParams5d)="CC"
upCC_5d=hyperGTest(upParams5d)
upCC1_5d=summary(upCC_5d)

ontology(downParams5d)="CC"
downCC_5d=hyperGTest(downParams5d)
downCC1_5d=summary(downCC_5d)

##
ontology(upParams5d)="MF"
upMF_5d=hyperGTest(upParams5d)
upMF1_5d=summary(upMF_5d)

ontology(downParams5d)="MF"
downMF_5d=hyperGTest(downParams5d)
downMF1_5d=summary(downMF_5d)
```


```{r}
#WKY 
resWKY$id<-rownames(resWKY)
resWKY$entrez<-mapIds(org.Rn.eg.db,key=resWKY$id,column="ENTREZID",keytype="SYMBOL")
resWKY5=as.data.frame(subset(resWKY,padj<0.05))
universeGenes=resWKY$entrez

selectGenesUpWKY=subset(resWKY5,resWKY5$log2FoldChange>0)$entrez
selectGenesDownWKY=subset(resWKY5,resWKY5$log2FoldChange<0)$entrez

upParamsWKY=new("GOHyperGParams",geneIds=selectGenesUpWKY,universeGeneIds=universeGenes,annotation="org.Rn.eg.db",ontology="BP",pvalueCutoff=0.05, conditional=FALSE,testDirection="over")
downParamsWKY=new("GOHyperGParams",geneIds=selectGenesDownWKY,universeGeneIds=universeGenes,annotation="org.Rn.eg.db",ontology="BP",pvalueCutoff=0.05, conditional=FALSE,testDirection="over")

upBP_WKY=hyperGTest(upParamsWKY)
upBP1_WKY=summary(upBP_WKY)

downBP_WKY=hyperGTest(downParamsWKY)
downBP1_WKY=summary(downBP_WKY)

##
ontology(upParamsWKY)="CC"
upCC_WKY=hyperGTest(upParamsWKY)
upCC1_WKY=summary(upCC_WKY)

ontology(downParamsWKY)="CC"
downCC_WKY=hyperGTest(downParamsWKY)
downCC1_WKY=summary(downCC_WKY)

##
ontology(upParamsWKY)="MF"
upMF_WKY=hyperGTest(upParamsWKY)
upMF1_WKY=summary(upMF_WKY)

ontology(downParamsWKY)="MF"
downMF_WKY=hyperGTest(downParamsWKY)
downMF1_WKY=summary(downMF_WKY)
```

```{r}
#SHR
resSHR$id<-rownames(resSHR)
resSHR$entrez<-mapIds(org.Rn.eg.db,key=resSHR$id,column="ENTREZID",keytype="SYMBOL")
resSHR5=as.data.frame(subset(resSHR,padj<0.05))
universeGenes=resSHR$entrez

selectGenesUpSHR=subset(resSHR5,resSHR5$log2FoldChange>0)$entrez
selectGenesDownSHR=subset(resSHR5,resSHR5$log2FoldChange<0)$entrez

upParamsSHR=new("GOHyperGParams",geneIds=selectGenesUpSHR,universeGeneIds=universeGenes,annotation="org.Rn.eg.db",ontology="BP",pvalueCutoff=0.05, conditional=FALSE,testDirection="over")
downParamsSHR=new("GOHyperGParams",geneIds=selectGenesDownSHR,universeGeneIds=universeGenes,annotation="org.Rn.eg.db",ontology="BP",pvalueCutoff=0.05, conditional=FALSE,testDirection="over")

upBP_SHR=hyperGTest(upParamsSHR)
upBP1_SHR=summary(upBP_SHR)

downBP_SHR=hyperGTest(downParamsSHR)
downBP1_SHR=summary(downBP_SHR)

##
ontology(upParamsSHR)="CC"
upCC_SHR=hyperGTest(upParamsSHR)
upCC1_SHR=summary(upCC_SHR)

ontology(downParamsSHR)="CC"
downCC_SHR=hyperGTest(downParamsSHR)
downCC1_SHR=summary(downCC_SHR)

##
ontology(upParamsSHR)="MF"
upMF_SHR=hyperGTest(upParamsSHR)
upMF1_SHR=summary(upMF_SHR)

ontology(downParamsSHR)="MF"
downMF_SHR=hyperGTest(downParamsSHR)
downMF1_SHR=summary(downMF_SHR)
```

##GO plot
```{r}
library('dplyr')
head(downMF1_WKY)
row.names(downMF1_WKY) <- downMF1_WKY$Term

downMF1_WKY$ratio=as.integer(downMF1_WKY$Count)/as.integer(downMF1_WKY$Size)
downMF1_WKY$Term <- factor(downMF1_WKY$Term, levels = downMF1_WKY$Term[order(-(downMF1_WKY$Pvalue))])
downMF1_WKY <- downMF1_WKY[1:20,]

ggplot(downMF1_WKY, aes(x=Count/Size, y=downMF1_WKY$Term, color=downMF1_WKY$Pvalue, size=downMF1_WKY$Count)) + geom_point() +  theme_classic() + scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab")+xlab("Rich Ratio")+labs(color="P value", size="Gene Number")+ylab("Downregulated GO Pathways (Molecular Function)") +ggtitle("WKY Joint Chon: 5-day-old vs 3-month-old")
```

```{r}
library('dplyr')
head(upMF1_SHR)
row.names(upMF1_SHR) <- upMF1_SHR$Term

upMF1_SHR$ratio=as.integer(upMF1_SHR$Count)/as.integer(upMF1_SHR$Size)
upMF1_SHR$Term <- factor(upMF1_SHR$Term, levels = upMF1_SHR$Term[order(-(upMF1_SHR$Pvalue))])
upMF1_SHR <- upMF1_SHR[1:20,]

ggplot(upMF1_SHR, aes(x=Count/Size, y=upMF1_SHR$Term, color=upMF1_SHR$Pvalue, size=upMF1_SHR$Count)) + geom_point() +  theme_classic() + scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab")+xlab("Rich Ratio")+labs(color="P value", size="Gene Number")+ylab("Upregulated GO Pathways (Molecular Function)") +ggtitle("SHR Joint Chon: 5-day-old vs 3-month-old")
```

```{r}
library('dplyr')
head(downBP1_SHR)
row.names(downBP1_SHR) <- downBP1_SHR$Term

downBP1_SHR$ratio=as.integer(downBP1_SHR$Count)/as.integer(downBP1_SHR$Size)
downBP1_SHR$Term <- factor(downBP1_SHR$Term, levels = downBP1_SHR$Term[order(-(downBP1_SHR$Pvalue))])

downBP1_SHR <- downBP1_SHR[1:20,]

ggplot(downBP1_SHR, aes(x=Count/Size, y=downBP1_SHR$Term, color=downBP1_SHR$Pvalue, size=downBP1_SHR$Count)) + geom_point() +  theme_classic() + scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab")+xlab("Rich Ratio")+labs(color="P value", size="Gene Number")+ylab("Downregulated GO Pathways (Biological Process)") +ggtitle("SHR Joint Chon: 5-day-old vs 3-month-old")
```

```{r}
library('dplyr')
head(downBP1_WKY)
row.names(downBP1_WKY) <- downBP1_WKY$Term

downBP1_WKY$ratio=as.integer(downBP1_WKY$Count)/as.integer(downBP1_WKY$Size)
downBP1_WKY$Term <- factor(downBP1_WKY$Term, levels = downBP1_WKY$Term[order(-(downBP1_WKY$Pvalue))])

downBP1_WKY <- downBP1_WKY[1:20,]

ggplot(downBP1_WKY, aes(x=Count/Size, y=downBP1_WKY$Term, color=downBP1_WKY$Pvalue, size=downBP1_WKY$Count)) + geom_point() +  theme_classic() + scale_color_gradient(low = "red2",  high = "mediumblue", space = "Lab")+xlab("Rich Ratio")+labs(color="P value", size="Gene Number")+ylab("Downregulated GO Pathways (Biological Process)") +ggtitle("WKY Joint Chon: 5-day-old vs 3-month-old")
```

#Combine normalized expression and differential analysis
```{r}
res3m1=as.data.frame(res3m)
rld3m1=as.data.frame(assay(rld3m))
rld3m1$id <- rownames(rld3m1)
res3m1$id <- rownames(res3m1)
cbind_3m<-inner_join(rld3m1,res3m1,by="id")
rownames(cbind_3m)<-cbind_3m$id

res5d1=as.data.frame(res5d)
rld5d1=as.data.frame(assay(rld5d))
rld5d1$id <- rownames(rld5d1)
res5d1$id <- rownames(res5d1)
cbind_5d<-inner_join(rld5d1,res5d1,by="id")
rownames(cbind_5d)<-cbind_5d$id

resWKY1=as.data.frame(resWKY)
rldWKY1=as.data.frame(assay(rldWKY))
rldWKY1$id <- rownames(rldWKY1)
resWKY1$id <- rownames(resWKY1)
WKY1<-inner_join(rldWKY1,resWKY1,by="id")
rownames(WKY1)<-WKY1$id

resSHR1=as.data.frame(resSHR)
rldSHR1=as.data.frame(assay(rldSHR))
rldSHR1$id <- rownames(rldSHR1)
resSHR1$id <- rownames(resSHR1)
SHR1<-inner_join(rldSHR1,resSHR1,by="id")
rownames(SHR1)<-SHR1$id

rldChon1=as.data.frame(assay(rldChon))
rldChon1$id <- rownames(rldChon1)
Chon1<-rldChon1
rownames(Chon1)<-Chon1$id

all_list<-list(cbind_3m,cbind_5d,WKY1,SHR1,Chon1)
names(all_list) <- c("3-month-old Joint: WKY vs SHR","5-day-old Joint: WKY vs SHR","WKYJoint: 5-day-old vs 3-month-old","SHR Joint: 5-day-old vs 3-month-old","Chondrocytes: WKY vs SHR")
```

#Select group for analysis (2 groups)
```{r}
#2 groups
x0 = all_list$`5-day-old Joint: WKY vs SHR`
#x0 = all_list$`3-month-old Joint: WKY vs SHR`
#x0 = all_list$`WKYJoint: 5-day-old vs 3-month-old`
#x0 = all_list$`SHR Joint: 5-day-old vs 3-month-old`

title_group=paste0(colnames(x0[1])," vs ",colnames(x0[4]))
title_group <- gsub("_A","",title_group)

x_rld=as.data.frame(dplyr::select(x0,1,2,3,4,5,6))
z=as.data.frame(dplyr::select(x0,8,9,10,11,12,13))

groups = c("WKY","WKY","WKY",'SHR','SHR','SHR')
#groups = c("5-day-old","5-day-old","5-day-old",'3-month-old','3-month-old','3-month-old')
```

```{r}
#Target pathways
kegg.sets.rn<- keggData$kg.sets
keggInt=as.data.frame(kegg.sets.rn$`rno04150 mTOR signaling pathway`)
colnames(keggInt)[1]<-"entrez"
keggInt$id=mapIds(org.Rn.eg.db,key=keggInt$entrez,column="SYMBOL",keytype="ENTREZID")
keggInt<-keggInt[-1]
ids<-keggInt
```

```{r}
###2 group
library(ComplexHeatmap)
library(dplyr)

#significance
z$id<-row.names(z)
z1=inner_join(x = ids, y =z , by = 'id')

z1$sig<-ifelse(z1$padj<0.0001,'****',
                    ifelse(z1$padj<0.001,'***',
                           ifelse(z1$padj<0.01,'**',
                                  ifelse(z1$padj<0.05,'*','ns'))))
z1$sig<-ifelse(z1$pval<0.05,z1$sig,'ns')

##significant only
#z1<-subset(z1,padj<0.05)
z1<-subset(z1,pvalue<0.05)

#expression
x=as.data.frame(x_rld)
x$id<-row.names(x)

x1=inner_join(x = ids, y =x , by = 'id')
rownames(x1)<-x1$id

x1=inner_join(x = x1, y =z1 , by = 'id')
x1=dplyr::select(x1,1,2,3,4,5,6,7)

rownames(x1)<-x1$id
x1<-x1[,-1]
x2=as.matrix(x1)

x3=x2

##draw heatmap

 c1<-c('cyan','#FF6EC7')
names(c1)<- unique(groups)
 c2<-c('yellow','orange','red','magenta','gray')
names(c2)<-c('*','**','***','****','ns')

ha=columnAnnotation("Groups"=groups,
                       col=list("Groups"=c1),  #if error, check group no.
                       show_annotation_name = TRUE,
                       # annotation_name_side="right",
                       annotation_name_gp = gpar(fontsize = 7),
                       show_legend=T,
                       simple_anno_size = unit(0.2, "cm"),
                       annotation_height=unit.c(unit(0.07, "cm")))

ha1 = rowAnnotation("Significance"=z1$sig,
                       col=list("Significance"=c2),
                       show_annotation_name = TRUE,
                       # annotation_name_side="right",
                       annotation_name_gp = gpar(fontsize = 7),
                       show_legend=T,
                       simple_anno_size = unit(0.2, "cm"),
                       annotation_height=unit.c(unit(0.07, "cm")))

ht<-Heatmap(x3,
                name='Normalized Expression',
                 top_annotation = ha,
                 cluster_rows = T,
                left_annotation = ha1,
                col=bluered(75),
                color_space = "RGB",
                cluster_columns = FALSE,
                row_order=NULL,
                show_column_names = T,
                show_row_names = T,
                row_names_gp = gpar(fontsize = 7),
                column_names_gp = gpar(fontsize = 7),
                gap = unit(3, "mm"),
                column_title_gp = gpar(fontsize = 7),
                width=unit(ncol(x3)*2.6/6, "cm"),
                height=unit(nrow(x3)*0.35, "cm"),
                show_heatmap_legend = T,
                heatmap_legend_param=list(labels_gp = gpar(fontsize = 7, fontface = "bold"),
                                          title_gp = gpar(fontsize = 7, fontface = "bold"))) 

 draw(ht)
```

#Overall analysis (4 groups)
```{r}
#group
x0 = all_list$`Chondrocytes: WKY vs SHR`

title_group="Chondrocytes WKY vs SHR"
x_rld=as.data.frame(dplyr::select(x0,1,2,3,4,5,6,7,8,9,10,11,12))

x_3m = all_list$`3-month-old Joint: WKY vs SHR`
z_3m=as.data.frame(dplyr::select(x_3m,8,9,10,11,12,13))
x_5d = all_list$`5-day-old Joint: WKY vs SHR`
z_5d=as.data.frame(dplyr::select(x_5d,8,9,10,11,12,13))
x_S = all_list$`SHR Joint: 5-day-old vs 3-month-old`
z_S=as.data.frame(dplyr::select(x_S,8,9,10,11,12,13))
x_W = all_list$`WKYJoint: 5-day-old vs 3-month-old`
z_W=as.data.frame(dplyr::select(x_W,8,9,10,11,12,13))

groups = c("5do WKY","5do WKY","5do WKY","3mo WKY","3mo WKY","3mo WKY",'5do SHR','5do SHR','5do SHR','3mo SHR','3mo SHR','3mo SHR')
```

```{r}
library(stringr)
library(dplyr)
#Target genes
ids1=read.csv('D:/WKY_vs_SHR/ids1.csv',header=1)
ids1<-mutate_all(ids1, .funs=str_to_title)

ids=as.data.frame(as.character(ids1$Insulin.like.Growth.Factor.Signalling))
colnames(ids)[1]<-"id"
```

```{r}
###4 groups
library(ComplexHeatmap)

library(gplots)
#expression
x=as.data.frame(x_rld)
x$id<-row.names(x)

x1=inner_join(x = ids, y =x , by = 'id')
rownames(x1)<-x1$id

#significance for individual comparison
z_3m$id<-row.names(z_3m)
z_5d$id<-row.names(z_5d)
z_S$id<-row.names(z_S)
z_W$id<-row.names(z_W)

z_5d<-subset(z_5d,z_5d$baseMean>0)
z_3m<-subset(z_3m,z_3m$baseMean>0)
z_S<-subset(z_S,z_S$baseMean>0)
z_W<-subset(z_W,z_W$baseMean>0)

z_3m$sig_3m<-ifelse(z_3m$padj<0.0001,'****',
                    ifelse(z_3m$padj<0.001,'***',
                           ifelse(z_3m$padj<0.01,'**',
                                  ifelse(z_3m$padj<0.05,'*','ns'))))
z_3m$sig_3m<-ifelse(z_3m$pval<0.05,z_3m$sig_3m,'ns')
z_5d$sig_5d<-ifelse(z_5d$padj<0.0001,'****',
                    ifelse(z_5d$padj<0.001,'***',
                           ifelse(z_5d$padj<0.01,'**',
                                  ifelse(z_5d$padj<0.05,'*','ns'))))
z_5d$sig_5d<-ifelse(z_5d$pval<0.05,z_5d$sig_5d,'ns')
z_W$sig_W<-ifelse(z_W$padj<0.0001,'****',
                    ifelse(z_W$padj<0.001,'***',
                           ifelse(z_W$padj<0.01,'**',
                                  ifelse(z_W$padj<0.05,'*','ns'))))
z_W$sig_W<-ifelse(z_W$pval<0.05,z_W$sig_W,'ns')
z_S$sig_S<-ifelse(z_S$padj<0.0001,'****',
                    ifelse(z_S$padj<0.001,'***',
                           ifelse(z_S$padj<0.01,'**',
                                  ifelse(z_S$padj<0.05,'*','ns'))))
z_S$sig_S<-ifelse(z_S$pval<0.05,z_S$sig_S,'ns')

x1=inner_join(x = x1, y =z_3m , by = 'id')
x1=inner_join(x = x1, y =z_5d , by = 'id')
x1=inner_join(x = x1, y =z_W , by = 'id')
x1=inner_join(x = x1, y =z_S , by = 'id')

##significant only
#x1<-subset(x1, padj.x<0.05 | padj.y<0.05 | padj.x.x<0.05 | padj.y.y<0.05)
#x1<-subset(x1, padj.x.x<0.05 | padj.y.y<0.05)    #time
#x1<-subset(x1, padj.x<0.05 | padj.y<0.05)   #strain

x2=dplyr::select(x1,1,2,3,4,5,6,7,8,9,10,11,12,13)

rownames(x2)<-x1$id
x2<-x2[,-1]
x2=as.matrix(x2)

x3=x2

##draw heatmap

 c1<-c('cyan','#FF6EC7','limegreen','purple')
names(c1)<- unique(groups)
 c2<-c('yellow','orange','red','magenta','gray')
names(c2)<-c('*','**','***','****','ns')

ha=columnAnnotation("Groups"=groups,
                       col=list("Groups"=c1),
                       show_annotation_name = TRUE,
                       # annotation_name_side="right",
                       annotation_name_gp = gpar(fontsize = 7),
                       show_legend=T,
                       simple_anno_size = unit(0.2, "cm"),
                       annotation_height=unit.c(unit(0.07, "cm")))

ha1 = rowAnnotation("3-month: WKY vs SHR"=x1$sig_3m,
                    "5-day: WKY vs SHR"=x1$sig_5d,
                    "WKY: 3-month vs 5-day"=x1$sig_W,
                    "SHR: 3-month vs 5-day"=x1$sig_S,
                       col=list("5-day: WKY vs SHR"=c2,"WKY: 3-month vs 5-day"=c2,"SHR: 3-month vs 5-day"=c2,"3-month: WKY vs SHR"=c2),
                       show_annotation_name = TRUE,
                       # annotation_name_side="right",
                       annotation_name_gp = gpar(fontsize = 7),
                       show_legend=T,
                       simple_anno_size = unit(0.2, "cm"),
                       annotation_height=unit.c(unit(0.07, "cm")))

ht<-Heatmap(x3,
                name='Normalized Expression',
                 top_annotation = ha,
                 cluster_rows = T,
                left_annotation = ha1,
                col=bluered(75),
                color_space = "RGB",
                cluster_columns = FALSE,
                row_order=NULL,
                show_column_names = T,
                show_row_names = T,
                row_names_gp = gpar(fontsize = 7),
                column_names_gp = gpar(fontsize = 7),
                gap = unit(3, "mm"),
                column_title_gp = gpar(fontsize = 7),
                width=unit(ncol(x3)*2.6/6, "cm"),
                height=unit(nrow(x3)*0.35, "cm"),
                show_heatmap_legend = T,
                heatmap_legend_param=list(labels_gp = gpar(fontsize = 7, fontface = "bold"),
                                          title_gp = gpar(fontsize = 7, fontface = "bold"))) 

 draw(ht)
```

##Boxplot
```{r}
gene_id="Igf1"

#Order (left to right): WKY 5d, WKY 3m, SHR 5d, SHR 3m

options(scipen=999)

pval12=z_W[gene_id,6] 
pval13=z_5d[gene_id,6] 
pval34=z_S[gene_id,6] 
pval24=z_3m[gene_id,6] 

group1="5do WKY"
group2="3mo WKY"
group3="5do SHR"
group4="3mo SHR"

all_rld<-x_rld

exprs1=all_rld[gene_id,1:3]
exprs1<- as.data.frame(t(exprs1))
colnames(exprs1)=c("Expression")
exprs1$Strain=c(group1)

exprs2=all_rld[gene_id,4:6]
exprs2<- as.data.frame(t(exprs2))
colnames(exprs2)=c("Expression")
exprs2$Strain=c(group2)

exprs3=all_rld[gene_id,7:9]
exprs3<- as.data.frame(t(exprs3))
colnames(exprs3)=c("Expression")
exprs3$Strain=c(group3)

exprs4=all_rld[gene_id,10:12]
exprs4<- as.data.frame(t(exprs4))
colnames(exprs4)=c("Expression")
exprs4$Strain=c(group4)

exprs0=rbind(exprs1,exprs2,exprs3,exprs4)

y_max=colMaxs(as.matrix(exprs0[,1]))+0.3
y_min=colMins(as.matrix(exprs0[,1]))-0.3

pval12_y=colMaxs(as.matrix(exprs0[,1]))*1.01
pval13_y=colMaxs(as.matrix(exprs0[,1]))*1.01+(y_max-y_min)*y_max/(8*y_max)
pval24_y=colMaxs(as.matrix(exprs0[,1]))*1.01+(y_max-y_min)*y_max/(16*y_max)
pval34_y=colMaxs(as.matrix(exprs0[,1]))*1.01

if(pval12<0.0001){
  df12<-data.frame(group1=group1, group2=group2, p.adj="p < 0.0001",y.position=pval12_y)
  } else if(pval12>0.1) {
  df12<-data.frame(group1=group1, group2=group2, p.adj="ns",y.position=pval12_y)
  } else if (pval12>0.0001 && pval12<0.1){ pval12<-round(pval12,4)
  df12<-data.frame(group1=group1, group2=group2,p.adj=paste0("p = ", as.character(format(pval12, scientific=F))), y.position=pval12_y)
  }
if(pval13<0.0001){
  df13<-data.frame(group1=group1, group2=group3, p.adj="p < 0.0001",y.position=pval13_y)
  } else if(pval13>0.1) {
  df13<-data.frame(group1=group1, group2=group3, p.adj="ns",y.position=pval13_y)
  } else if (pval13>0.0001 && pval13<0.1)  { pval13<-round(pval13,4)
  df13<-data.frame(group1=group1, group2=group3,p.adj=paste0("p = ", as.character(format(pval13, scientific=F))), y.position=pval13_y)
  }
if(pval24<0.0001){
  df24<-data.frame(group1=group2, group2=group4, p.adj="p < 0.0001",y.position=pval24_y)
  } else if (pval24>0.1) {
  df24<-data.frame(group1=group2, group2=group4, p.adj="ns",y.position=pval24_y)
  } else if (pval24>0.0001 && pval24<0.1) { pval24<-round(pval24,4)
  df24<-data.frame(group1=group2, group2=group4,p.adj=paste0("p = ", as.character(format(pval24, scientific=F))), y.position=pval24_y)
  }
if(pval34<0.0001){
  df34<-data.frame(group1=group3, group2=group4, p.adj="p < 0.0001",y.position=pval34_y)
 } else if (pval34>0.1) {
  df34<-data.frame(group1=group3, group2=group4, p.adj="ns",y.position=pval34_y)
  } else if (pval34>0.0001 && pval34<0.1) { pval34<-round(pval34,4)
  df34<-data.frame(group1=group3, group2=group4,p.adj=paste0("p = ", as.character(format(pval34, scientific=F))), y.position=pval34_y)
  }

ggplot(exprs0, aes(x=fct_inorder(Strain), y=Expression,color=fct_inorder(Strain))) +
  scale_color_manual(values=c("blue","red","green","purple")) +
  geom_boxplot(width=0.5,outlier.colour="yellow", outlier.shape=8,outlier.size=2,show.legend=FALSE) +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=1,show.legend=FALSE) + 
  labs(title=gene_id) + xlab("") + ylab("Normalized Expression") +
  theme_classic() + 
  theme(text = element_text(color="black",size=12,family = "sans"))+
  theme(plot.title = element_text(size=12,hjust=0.5,face="bold"))+
  theme(axis.title.y = element_text(size=12.5,face="bold"))+
  theme(axis.text.x = element_text(angle = 45, hjust=1,face="bold"))+
  theme(axis.text.y = element_text(face="bold"))+
  theme(plot.background = element_rect(fill = "transparent",colour = NA_character_))+
  theme(panel.background = element_rect(fill = "white"))+
  ylim(y_min,y_max)+ scale_y_continuous(n.breaks=7)+
  stat_pvalue_manual(df12, label = "p.adj",size=2.5,vjust=-0.05, bracket.shorten=0.03)+
  stat_pvalue_manual(df13, label = "p.adj",size=2.5,vjust=-0.05,bracket.shorten=0.03)+
  stat_pvalue_manual(df24, label = "p.adj",size=2.5,vjust=-0.05,bracket.shorten=0.03)+
  stat_pvalue_manual(df34, label = "p.adj",size=2.5,vjust=-0.05,bracket.shorten=0.03)
```


##Correlation matrix
```{r}
library(rstatix)
library(ggcorrplot)

ids1=read.csv('D:/WKY_vs_SHR/ids1.csv',header=1)
ids1<-mutate_all(ids1, .funs=str_to_title)
ids=as.data.frame(as.character(ids1$Insulin.like.Growth.Factor.Signalling))
ids<-subset(ids,ids[,1]!="")
gene_ids_1<-ids

ids1=read.csv('D:/WKY_vs_SHR/ids1.csv',header=1)
ids1<-mutate_all(ids1, .funs=str_to_title)
ids=as.data.frame(as.character(ids1$glycolysis.MARKERS))
ids<-subset(ids,ids[,1]!="")
gene_ids_2<-ids

matrix_1<-as.data.frame(matrix(ncol=ncol(x_rld)))
colnames(matrix_1)<-colnames(x_rld)
for (i in 1:nrow(gene_ids_1)){
  gene_no_1<-gene_ids_1[i,] 
  exprs_1=x_rld[gene_no_1,]
  matrix_1=rbind(exprs_1,matrix_1)
}
matrix_1<-na.omit(matrix_1)
matrix_1<-t(matrix_1)

matrix_2<-as.data.frame(matrix(ncol=ncol(x_rld)))
colnames(matrix_2)<-colnames(x_rld)
for (i in 1:nrow(gene_ids_2)){
  gene_no_2<-gene_ids_2[i,] 
  exprs_2=x_rld[gene_no_2,]
  matrix_2=rbind(exprs_2,matrix_2)
}
matrix_2<-na.omit(matrix_2)
matrix_2<-t(matrix_2)
matrix_corr <-cor(x=matrix_1,y=matrix_2, method = "pearson")

matrix_corr2<-cor(cbind(matrix_1,matrix_2))
p.mat <- cor_pmat(cbind(matrix_1,matrix_2))
p.mat_sig <- p.mat<0.05
p.mat_5<-ifelse(p.mat_sig,p.mat,"ns")

ggcorrplot(matrix_corr2, p.mat = p.mat, legend.title ="Pearson r", outline.color = "black" , 
           type="lower",insig = "blank",colors = c("blue2", "white", "red2"),sig.level = 0.1, 
           tl.cex = 8 , lab=TRUE, digits = 2, lab_size = 2,method = "square") 
```